---
title: "Time Series Decomposition"
author: "Dan Smilowitz"
date: "September 5, 2017"
output: 
  ioslides_presentation: 
    widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE, comment=NA,
                      fig.height = 5, fig.width = 10, fig.align = "center")
library(fma)
library(tidyverse)
```

## Background

Time series can exhibit variety of patterns, which can be split into components representing different patterns

Extracting patterns & components yields two benefits:

  - Understand series better
  - Improve forecasts

## Time Series Patterns
  
Three types of patterns:

  1.  **Trend:** long-term increase or decrease in data
    - May change direction
  2.  **Seasonal:** influenced by seasonal factors
    - Always of fixed, known period
  3.  **Cyclic:** rises and falls *not* of fixed period
  
## Time Series Patterns

```{r pattern-examples}
theme_set(theme_light())
gridExtra::grid.arrange(
  autoplot(hsales, xlab = "Year", ylab = "Housing sales (M)") +
    ggtitle("Monthly housing sales", "Seasonality & cyclical"),
  autoplot(ustreas, xlab = "Day", ylab = "Contracts") +
    ggtitle("US treasury bill contracts", "Downward trend"),
  autoplot(elec, xlab = "Year", ylab = "Production") +
    ggtitle("Australian monthly electricity production", "Seasonality & upward trend"),
  autoplot(diff(dj), xlab = "Day", ylab = "Daily change") +
    ggtitle("Daily change in Dow Jones index", "No strong patterns"),
  nrow = 2
)
```


## Time Series Components
A series $y_t$ has 3 components:

  - seasonal component $S_t$ (fixed-period variation)
  - trend-cycle component $T_t$ (long-term trend & non-fixed cycle)
  - remainder component $E_t$ (anything else in time series)

<div class="columns-2">

  
  **Additive Model:**
  $$y_t = S_t + T_t + E_t$$
  
  Seasonal fluctuation magnitude & trend-cycle variation are not proportional to level of time series
  
  <br />
  <br />
  <br />
  <br />
  
  
  **Multiplicative Model:**
  $$y_t = S_t \times T_t \times E_t$$
  
  Seasonal fluctuation magnitude & trend-cycle variation varies with level
  
  *Transformed:* $\log y_t = \log S_t + \log T_t + \log E_t$
  
</div>



## Seasonally Adjusted Data
The data with the seasonal component removed ($y_t - S_t$ or $y_t / S_t$) is the **seasonally adjusted data**.

  > - Useful if seasonal variation is not pattern of interest
  > - Comprises remainder and trend-cycle components
  > - Commonly used for economic data to study non-seasonal variation


## Moving Averages
Moving average of order $m$ ($m$-MA):
$$\hat{T}_t = \frac{1}{m} \sum_{j = -k}^k y_{t+j}$$
  
  > - Averages values $y_j$ within $k$ periods of $t$
  > - Eliminates some randomness, estimating trend-cycle
  > - Captures trend without minor fluctuation
  > - Increasing order increases smoothness


## Moving Averages | MA Order

```{r ma-order}
data(elecsales, package = "fpp")
data_frame(Year = time(elecsales),
           GWh = as.numeric(elecsales),
           `3-MA` = ma(elecsales, order = 3),
           `5-MA` = ma(elecsales, order = 5),
           `7-MA` = ma(elecsales, order = 7),
           `9-MA` = ma(elecsales, order = 9)) %>% 
  gather(Order, Value, -c(Year, GWh)) %>% 
  ggplot(aes(x = Year)) + geom_line(aes(y = GWh)) +
  geom_line(aes(y = Value, col = Order), lwd = 1, show.legend = FALSE) +
  facet_wrap(~ Order, nrow = 2) +
  ggtitle("Residential electricity sales", "Data vs. moving averages of varying order")
```


## Moving Averages | Averages of Averages

  - The moving average of a moving average gives a smoother trend estimate
  - Even-order MA of even-order MA will return a symmetric MA
    - *i.e.* takes same number of points before and after $t$
  - Notation $m_2 \times m_1$-MA
  
### Estimating Trend with Seasonal Data
Taking $2 \times m$-MA yields weighted MA of period $m+1$ with first and last observations having half-weight.  This can be used to view the trend absent seasonality


## Moving Averages

```{r elec-2x12}
autoplot(elec, xlab = "Year", ylab = "GWh") +
  ggtitle("Residential electricity sales", "Data and 2x12-MA") +
  geom_line(aes(t, ma), data_frame(t = time(elec), ma = ma(elec, 12)), col = 2)
```

## Classical Decomposition
<div class="columns-2">
  **Additive decomposition:**

  1. Compute trend cycle component $\hat{T}_t$
    - $2 \times m$-MA if $m$ is even
    - $m$-MA if $m$ is odd
  2. Calculate de-trended series $y_t - \hat{T}_t$
  3. Calculate $\hat{S}_t$ from de-trended values
    + Average values for each period
    + Adjust values so sum is zero
  4. Remainder is $\hat{E}_t = y_t - \hat{T}_t - \hat{S}_t$
  
  `decompose(x, type = "a")`
  
  <br />
  
  **Multiplicative decomposition:**

  1. Compute trend cycle component $\hat{T}_t$
    - $2 \times m$-MA if $m$ is even
    - $m$-MA if $m$ is odd
  2. Calculate de-trended series $y_t / \hat{T}_t$
  3. Calculate $\hat{S}_t$ from de-trended values
    + Average values for each period
    + Adjust values so sum is $m$
  4. Remainder is $\hat{E}_t = y_t / \left( \hat{T}_t - \hat{S}_t \right)$
  
  `decompose(x, type = "m")`
  
</div>

## Classical Decomposition | Shortcomings

  - Trend estimate unavailable for first & last $m/2$ periods
    - Therefore no remainder $\rightarrow$ all seasonality
  - Assume seasonal component repeats from year to year
    - Unable to capture changes in seasonality over time
  - Unable to detect unusual patterns

### X-12-ARIMA
Method created to address shortcomings --- 16-step process detailed in section *6/4* (pages 162-163).  Requires proprietary software; no R package for performing this decomposition.

## STL Decomposition
**S**easonal and **T**rend decomposition using **L**oess (method for estimating nonlinear relationships) presents strengths over other methods:

  - Handle any type of seasonality
  - Seasonal component allowed to change over time
  - User-controlled trend-cycle smoothness
  - Robust to unusual observations
  
STL can only be used with additive decomposition -- must log-transform multiplicative models

`stl(x, s.window, t.window, ...)`

## STL Decomposition
```{r stl-periodic}
library(highcharter)
hchart(stl(elec, "periodic")) %>%
  hc_title(text = "STL Decomposition of Australian Electricity Usage") %>% 
  hc_subtitle(text = "Periodic seasonal component")
```

## STL Decomposition
```{r stl-seasonality-change}
hchart(stl(elec, s.window = 13, robust = TRUE)) %>%
  hc_title(text = "STL Decomposition of Australian Electricity Usage") %>% 
  hc_subtitle(text = "Changing seasonal component (s.window = 13); Robust to outliers")
```


## Foreasting with Decomposition
A decomposed time series is written as:

  - $y_t = \hat{S}_t + \hat{A}_t$ with $\hat{A}_t = \hat{T}_t + \hat{E}_t$ (additive)
  - $y_t = \hat{S}_t  \hat{A}_t$ with $\hat{A}_t = \hat{T}_t  \hat{E}_t$ (multiplicative)

The two components $\hat{S}_t$ & $\hat{A}_t$ are forecast separately:

  - $\hat{S}_t$ assumed unchanging
    - Prior year's value taken (seasonal naive)
  - $\hat{A}_t$ forecast using non-seasonal method
    - Naive
    - Random walk with drift
    - ARIMA

## Foreasting with Decomposition
```{r forecast}
data(elecequip, package = "fpp")
elecequip %>% 
  stl(s.window = "periodic", t.window = 15, robust = TRUE) %>% 
  forecast(method = "rwdrift") %>% 
  autoplot() +
  labs(title = "Forecast of electric equipment manufacturing",
       subtitle = "Forecasts from STL + Random walk with drift",
       x = "Year", y = "New orders index")
```


# Thank You
### Slides: https://rpubs.com/dsmilo/DATA624-Lecture
### Source code: https://github.com/dsmilo/DATA624
